---
title: "Writing reports in Rmarkdown (HTML version)"
author: "Daniel Hammarström"
date: "7 11 2021"
output: 
  bookdown::html_document2: default
bibliography: resources/references.bib
csl: resources/vancouver.csl
---

# Introduction

-   R Markdown documents are source documents combining code and text into reproducible reports.
-   R Markdown can be converted to multiple output formats including HTML, word and PDF. These outputs have several other options to produce presentations, books or simple reports.
-   Reports can be customized to fit requirements for your thesis or paper.

# Basics

## YAML and bibliography

-   The YAML field lets you specify options for your output file. Not all fields will be read by your output. A simple example that produces the `title`, `author`, `date` and sets the output format of this document (`output`).

```{r, eval = FALSE}

---
title: "Writing reports in Rmarkdown (HTML version)"
author: "Daniel Hammarström"
date: "7 11 2021"
output: html_document
---
 
```

-   The YAML field is used to specify the bibliography. This is a file containing references in a format compatible with the reference engine, usually in bibtex format.

-   To add a bibliography, specify `bibliography:` with the path to your bibliography.

```{r, eval = FALSE}



---
title: "Writing reports in Rmarkdown (HTML version)"
author: "Daniel Hammarström"
date: "7 11 2021"
output: html_document
bibliography: resources/references.bib
---

```

-   In the file `references.bib` a bibtex reference is created such as:

```{r, eval = FALSE}

@article{hammarström2020,
	title = {Benefits of higher resistance-training volume are related to ribosome biogenesis},
	author = {{Hammarström}, {Daniel} and {Øfsteng}, {Sjur} and {Koll}, {Lise} and {Hanestadhaugen}, {Marita} and {Hollan}, {Ivana} and {Apró}, {William} and {Whist}, {Jon Elling} and {Blomstrand}, {Eva} and {Rønnestad}, {Bent R.} and {Ellefsen}, {Stian}},
	year = {2020},
	date = {2020},
	journal = {The Journal of physiology},
	pages = {543--565},
	volume = {598},
	number = {3},
	doi = {10.1113/JP278455},
	url = {https://www.ncbi.nlm.nih.gov/pubmed/31813190 https://physoc.onlinelibrary.wiley.com/doi/full/10.1113/JP278455},
	note = {Edition: 2020/01/15},
	langid = {eng}
}


```

-   If you have Zotero installed, adding references is as easy as opening the visual markdown editor and add citation.

-   In the final report, citations will look e.g. like this [@hammarström2020;@wen2016]. This is the result of writing `[@hammarström2020;@wen2016]` in your markdown document.
-   Your references must be in your `references.bib` file. If using the plug-in, new references will be updated to the file as you reference.

## Changing the citation style

- Citation styles exists in many different shapes (e.g. (Wen 2016), [2], (2), ^2^)
- To change the citation style in an R Markdown document you may add a custom csl-file. Citation style language (CSL) is, as the name implies a language for citation styles. Find more information [here](https://citationstyles.org/authors/).
- Citation styles can be downloaded from the github folder [https://github.com/citation-style-language/styles](https://github.com/citation-style-language/styles). Copy the code from github into a new file and store it in your resources folder. Add `csl: resources/vancouver.csl` to your YAML:

```{r, eval = FALSE}
---
title: "Writing reports in Rmarkdown (HTML version)"
author: "Daniel Hammarström"
date: "7 11 2021"
output: 
  bookdown::html_document2: default
bibliography: resources/references.bib
csl: resources/vancouver.csl
---
  
```



## Cross-referencing

-   You can use cross-referencing inside documents to refer to figures, tables and headlines.
-   Cross-referencing works when setting the output to `html_document2` in the YAML field, this output is specified in the `bookdown` package so the YAML field will havce to look like this:

```{r eval = FALSE}

---
title: "Writing reports in Rmarkdown (HTML version)"
author: "Daniel Hammarström"
date: "7 11 2021"
output: 
  bookdown::html_document2: default
bibliography: resources/references.bib
---

```


-   To cross-reference a figure, you also have to name the code-chunk producing the figure, e.g. `{r figure-1}`.
-   You may now reference the figure using `\@ref(fig:figure-1)`. The reference will appear as a number (see Figure \@ref(fig:figure-1)) and the figure will be automatically numbered.


```{r figure-1, warning=FALSE, message = FALSE, fig.cap = "An example caption"}
library(tidyverse);library(exscidata)
data(dxadata)

dxadata %>%
  ggplot(aes(height, lean.legs)) + geom_point()


```

-   Similarly, tables can also be referenced but the need another type `\@ref(table:table-1)`
-   Headers can be referenced using e.g. `[see References](#refs)` where `#refs` is a id that I have put after the References heading below. This will produce (see [References](#refs)).



## Including HTML

-   Since the report engine converts your document to HTML you may add HTML code. E.g. if you would like to embed a youtube clip it is as simple as copying the code from youtube to your document.

```{r, eval = FALSE}

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/_DhOMdcbjuU" frameborder="0" allowfullscreen></iframe>

```

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/_DhOMdcbjuU" frameborder="0" allowfullscreen></iframe>

# Exercises

1. Define a new R project and create a R Markdown file.
2. Organize the project folder and add a file containing bibtex citations, write some text and add a citation.
3. Explore the CSL github repository and change your csl format.
3. Add a figure and a table (`kable`) to the document, cross-reference both.




# References {#refs}
