---
title: "Correlated data"
date: "2021-11-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Alternatives when analyzing correlated data

When we have more observations from each experimental unit (participant, family, classroom etc.) we have correlated data. This needs to be accounted for in the statistica model. In the `dxadata` and `strengthvolume` data sets multiple measurements are collected from each participants. 

Prepare the data:

```{r}
library(tidyverse); library(exscidata)
data("dxadata")


dat <- dxadata %>%
  select(participant, time,sex, multiple, single, include,
         lean.left_leg, lean.right_leg) %>%
  pivot_longer(names_to = "leg", 
               values_to = "lean.mass", 
               cols = lean.left_leg:lean.right_leg) %>% 
  mutate(leg = if_else(leg == "lean.left_leg", "L", "R"), 
         sets = if_else(multiple == leg, "multiple", "single")) %>%
  filter(include == "incl") %>%
  
  select(participant, time, sex, sets, leg, lean.mass) %>%
  
  pivot_wider(names_from = time, 
              values_from = lean.mass)  %>%
  mutate(lbm.change = post - pre, 
         pre.mc = pre - mean(pre)) %>%
 
  print()
```

To fit models with correlated data we need to use mixed-effects models that account for the structure of the data set. The package `lme4` has such functions (`lmer`). The package `lmerTest` provides the same function but with the addition of p-values.

Using a ordinary linera model (`lm`) when we have correlation in the data will in some situations lead to false negative results. 

```{r}
library(lme4); library(lmerTest)


m0 <- lm(post ~ pre +sex + sets, data = dat)
m1 <- lmerTest::lmer(post ~ pre + sex + sets + (1|participant), data = dat)
m2 <- lme4::lmer(post ~ pre + sex + sets + (1|participant), data = dat)


plot(m2)


summary(m0)
summary(m2)
summary(m1)
```

Inference from `lme4` (without p-values) can be made based on confidence intervals. 

```{r}
confint(m2)

```






