---
title: "Untitled"
author: "Daniel Hammarström"
date: "8 11 2021"
output: pdf_document
toc: true
lot: true
lof: true
editor_options: 
  chunk_output_type: console
bibliography: resources/references.bib
---

# Introduction

-   PDF output requires a working LaTeX installation on your computer. The easiest way to do this installation is by installing `tinytex`.

```{r, eval = FALSE}

install.packages('tinytex')
tinytex::install_tinytex()  # install TinyTeX

```

-   By selecting PDF as output, R Markdown documents will be converted trough LaTeX, a special type setting engine. This makes the output very flexible. Flexibility comes with a price, complexity.
-   The PDF output should be considered when you want a nice format that is easy to read.
-   Similar to other outputs you can add components such as figures[@wickham2016], tables and citations with a high degree of customization.

# Table of content, list of table and figures

-   By adding `toc: true`, `lot: true` and `lof: true` to the YAML you will get a table of content and list of tables and figures.

## A subheading
-   Using subheadings will produce hierarchy in your `toc` 

# References

Sometimes you might want to add references in the middle of the document, this is supported in all outputs by specifying `<div id="refs"> </div>`

<div id="refs"> </div>

# Figures and tables after references

```{r table-1, echo = FALSE, warning=FALSE, message=FALSE}

library(tidyverse); library(exscidata); library(knitr); data(dxadata)

dxadata %>%
  filter(time == "pre") %>%
  group_by(sex) %>%
  summarise(age = paste0(round(mean(age), 1), " (", round(sd(age), 1), ")")) %>%
  kable(col.names = c("Sex", "Age [Mean (SD)]"), 
        booktabs = TRUE, 
        caption = "Participant characteristics") 

```



```{r figure-1, echo = FALSE, warning=FALSE, message=FALSE, fig.cap = "An example figure"}

library(tidyverse); library(exscidata); library(knitr); data(dxadata)

dxadata %>%
  filter(time == "pre") %>%
  ggplot(aes(age, fill = sex)) +  geom_histogram()
  

```


# Possible issues with R Markdown and PDF outputs

- To much content may over-flow your pages, if your figures are too big etc. May need customization
- The LaTeX engine wont work with special characters. Switching to another LaTeX engine might help:

```{r, eval = FALSE}

output:
  pdf_document:
    latex_engine: xelatex

```


