---
title: "Creating nice figures"
date: "2021-11-25"
output: html_document
editor_options: 
  chunk_output_type: console
---


# The basic plot 

Before plotting it is a good idea to
- Reduce the data set to what you want to plot (e.g. mean and standard deviation from a specific exercise and time point)
- Change levels of factors (e.g. time).
- Decide on a basic setup, what to plot on x and y axis. Chose a `geom_`


```{r}

library(tidyverse); library(exscidata); data("strengthvolume")


strengthvolume %>%
  filter(exercise == "legext", time %in% c("pre", "post")) %>%
  group_by(sets, time) %>%
  summarise(m = mean(load, na.rm = TRUE), 
            s = sd(load, na.rm = TRUE)) %>%
  mutate(time = factor(time, levels = c("pre", "post"))) %>%
  ggplot(aes(time, m)) + geom_point()



```


# Adding multiple geoms, points and errorbars

- Error bars can be used to display the variation of the data. `geom_errobar`
- Points and errorbars will overlap, we need to dodge the geoms using 
- Add information on grouping to be able to "dodge" the geoms.

```{r}

strengthvolume %>%
  filter(exercise == "legext", time %in% c("pre", "post")) %>%
  group_by(sets, time) %>%
  summarise(m = mean(load, na.rm = TRUE), 
            s = sd(load, na.rm = TRUE)) %>%
  mutate(time = factor(time, levels = c("pre", "post"))) %>%
  ggplot(aes(time, m, color = sets)) + 
  
  geom_errorbar(aes(ymin = m - s, ymax = m + s), 
                position = position_dodge(width = 0.2), 
                width = 0.2) +
  geom_point(position = position_dodge(width = 0.2))


```

## Is this a suitable graph?

- Does the graph show the comparison we are interested in?
- Can we highlight the difference?
- Could we add a graph?


# Adding axes titles/labels

- Label axes and other aestethics using the `labs()` function
- To change labels of categories, add information to the data set.


```{r}

strengthvolume %>%
  filter(exercise == "legext", time %in% c("pre", "post")) %>%
  group_by(sets, time) %>%
  summarise(m = mean(load, na.rm = TRUE), 
            s = sd(load, na.rm = TRUE)) %>%
  mutate(time = factor(time, levels = c("pre", "post"), labels = c("Pre-training", "Post-training")), 
         sets = factor(sets, levels = c("single", "multiple"), labels = c("Single-set", "Multiple-set"))) %>%
  ggplot(aes(time, m, color = sets)) + 
  
  geom_errorbar(aes(ymin = m - s, ymax = m + s), 
                position = position_dodge(width = 0.2), 
                width = 0.2) +
  geom_point(position = position_dodge(width = 0.2)) +
  
  labs(x = "Time-point", 
       y = "Average load (kg)", 
       color = "Sets")


```


# Changing theme components


- Use a pre-built theme to get started (e.g. `theme_bw`)
- Change legend positions using the `theme()` function (see ?theme).


```{r}

strengthvolume %>%
  filter(exercise == "legext", time %in% c("pre", "post")) %>%
  group_by(sets, time) %>%
  summarise(m = mean(load, na.rm = TRUE), 
            s = sd(load, na.rm = TRUE)) %>%
  mutate(time = factor(time, levels = c("pre", "post"), labels = c("Pre-training", "Post-training")), 
         sets = factor(sets, levels = c("single", "multiple"), labels = c("Single-set", "Multiple-set"))) %>%
  ggplot(aes(time, m, color = sets)) + 
  
  geom_errorbar(aes(ymin = m - s, ymax = m + s), 
                position = position_dodge(width = 0.2), 
                width = 0.2) +
  geom_point(position = position_dodge(width = 0.2)) +
  
  labs(x = "Time-point", 
       y = "Average load (kg)", 
       color = "Sets") +
  
  theme_bw() + 
  
  theme(legend.position = c(0.1, 0.9))
  
```








