---
title: "Analyzing trials - Workshop 1"
author: "Daniel Hammarström"
date: "2 11 2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Interpreting the linear model

- A regression model can be used to assess the differences between treatments in a randomized controlled trial (RCT).
- In the simple case, this can be a comparison of change-scores


```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
data.frame(Time = c("Pre", "Post", "Pre", "Post"), outcome = c(0,1, 0,1)) %>%
        mutate(Time = factor(Time, levels = c("Pre", "Post"))) %>%
        ggplot(aes(Time, outcome)) + 
        theme_classic() + 
        ylab("Outcome") +
        theme(axis.text.y = element_blank()) +
        geom_segment(aes(x = Time, xend = 2, 
                         y = c(0.5,0.5, 0.5,1.5), 
                         yend = c(0.5,0.5, 1.5,1.5))) + 
        scale_y_continuous(limits = c(0,2)) + 
        annotate(geom = "text", x = 2.25, y = 1.5, label = "Treatment") +
        annotate(geom = "text", x = 2.25, y = 0.5, label = "Control") 


```


- The alternative hypothesis is that the two groups are different in terms of change scores.

$$ \Delta Y_{Treatment}-\Delta Y_{Control} \neq 0 $$

- The null-hypothesis that we use in our statistical models is 

$$ \Delta Y_{Treatment}-\Delta Y_{Control} = 0 $$

- In the simple case, this equals a *t*-test.

`t.test(delta ~ group, paired = FALSE)`

- A similar scenario is a comparison between two treatments

```{r, echo=FALSE}
data.frame(Time = c("Pre", "Post", "Pre", "Post"), outcome = c(0,1, 0,1)) %>%
        mutate(Time = factor(Time, levels = c("Pre", "Post"))) %>%
        ggplot(aes(Time, outcome)) + 
        theme_classic() + 
        ylab("Outcome") +
        theme(axis.text.y = element_blank()) +
        geom_segment(aes(x = Time, xend = 2, 
                         y = c(0.5,0.5, 0.5,1.5), 
                         yend = c(1.1,0.5, 1.5,1.5))) + 
        scale_y_continuous(limits = c(0,2)) + 
        annotate(geom = "text", x = 2.25, y = 1.5, label = "A") +
        annotate(geom = "text", x = 2.25, y = 1.1, label = "B") 

```

$$ \Delta Y_{A}-\Delta Y_{B} = 0 $$

- The `tenthirty` data set has similar data as in the above scenario
- We want to know what treatment to recommend for 1RM improvements. 




## Ten-thirty: Exercises

- Using the `tenthirty` data, use filter to keep time-point "pre" and "post" for the leg-press exercise.
- Make a plot of the averages of each group.
- Calculate a change score as `post` - `pre`.
- Make a plot of the average change score for each group.
- Fit a regression model (`change ~ group`) and interpret the results.


### Questions:

1. Concerning the sample: Which group has the largest improvement i leg-press 1RM? 
2. Concerning inference to the population: Which method do you recommend for improving 1RM leg-press?


### Solutions

Create a plot of averages over time:

```{r}
library(tidyverse); library(exscidata)
data("tenthirty")

# Save the data set in a filtered version
tenthirty_reduced <- tenthirty %>%
  filter(time %in% c("pre", "post"), 
         exercise == "legpress", 
         !is.na(load)) %>% 
  print()

# Create a line plot with one line per group
tenthirty_reduced %>%  
  group_by(time, group) %>%
  summarise(m = mean(load)) %>%
  mutate(time = factor(time, levels = c("pre", "post"))) %>%
  ggplot(aes(time, m, color = group, 
             group = group)) + geom_line()


# Create a plot of average changes per group
tenthirty_reduced %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre) %>%
  group_by(group) %>%
  summarise(mean_change = mean(change, na.rm = TRUE)) %>%
  ggplot(aes(group, mean_change, fill = group)) + 
  geom_bar(stat = "identity", width = 0.3)
  

# Calculate the average change per group, and difference between groups
tenthirty_reduced %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre) %>%
  group_by(group) %>%
  summarise(mean_change = mean(change, na.rm = TRUE)) %>%
  pivot_wider(names_from = group, 
              values_from = mean_change) %>%
  mutate(diff = RM30 - RM10) %>%
  print()


# Calculate change score and save new data set
dat <- tenthirty_reduced %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre) %>%
  print()


# Fit model with change scores explained by group
m1 <- lm(change ~ group, data = dat)

# 
summary(m1)



```



```{r, echo = TRUE, fig.cap = "Group averages over time"}
library(exscidata)

tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  mutate(time = factor(time, levels = c("pre", "post"))) %>%
  group_by(time, group) %>%
  summarise(m = mean(load, na.rm = TRUE)) %>%
  ggplot(aes(time, m, group = group, color = group)) + geom_line() +
  labs(x = "Time", y = "Legpress performance (kg)", color = "Group") +
          theme_classic()
```



## From t-test to ANCOVA

- Above we used the simple comparison of change score. A more appropriate model to study outcomes of a RCT is an analysis of co-variance (ANCOVA)
- In an ANCOVA change-scores or post values are compared and the estimates tells you the adjusted differences between groups if they had the exact same starting point.

### ANCOVA exercise

- Fit an ANCOVA model with change scores as the dependent variable
- Fit an ANCOVA model with the post values as the dependent variable
- Pre-training values and group should be covariates in both models.

### Questions

1. Concerning inference to the population: Which method do you recommend for improving 1RM leg-press?
2. In what ways are the models different?


### Soultions

```{r}

# Create a data set with change

dat <- tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre, 
         pre = pre - mean(pre)) %>% # Mean center pre-values
  print()


# Fit models 

m1 <- lm(post ~ pre + group, data = dat)

m2 <- lm(change ~ pre + group, data = dat)


summary(m1)
summary(m2)


m3 <- aov(post ~ pre + group, data = dat)

summary(m3)
anova(m1)


```







```{r, echo = FALSE}

m1 <- tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  pivot_wider(names_from = "time", values_from = "load") %>%
  mutate(change = post - pre) %>%
  lm(post ~ pre + group, data = .) 


tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
    pivot_wider(names_from = "time", values_from = "load") %>%
  ggplot(aes(pre, post, color = group)) + geom_point() + 
  scale_color_manual(values = c("red", "blue")) +
  geom_abline(intercept = coef(m1)[1], slope = coef(m1)[2], color = "red")  + 
  geom_abline(intercept = coef(m1)[1] + coef(m1)[3], slope = coef(m1)[2], color = "blue") + 
  labs(title = "Estimates from an ANCOVA model", 
       subtitle = "post ~ pre + group", 
       color = "Group") + theme_classic()



m2 <- tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  pivot_wider(names_from = "time", values_from = "load") %>%
  mutate(change = post - pre) %>%
  lm(change ~ pre + group, data = .) 


tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
    pivot_wider(names_from = "time", values_from = "load") %>%
    mutate(change = post - pre) %>%
  ggplot(aes(pre, change, color = group)) + geom_point() + 
  scale_color_manual(values = c("red", "blue")) +
  geom_abline(intercept = coef(m2)[1], slope = coef(m2)[2], color = "red")  + 
  geom_abline(intercept = coef(m2)[1] + coef(m2)[3], slope = coef(m2)[2], color = "blue") + 
  labs(title = "Estimates from an ANCOVA model", 
       subtitle = "change ~ pre + group", 
       color = "Group") + theme_classic()

```




## Hypertrophy and graded protein dose

- Read Haun et al. and find the main purpose/aim/hypothesis of the study.
- How do they motivate their purpose/aim/hypothesis?
- What is the statistical test used test the main hypotheses?



## Make it simple!

Design an analysis of the data from Haun et al.: 
  - include only two time-points (T1 and T3)
  - Select one meassure of muscle hypertrophy
  - Select a test for a null-hypothesis significance test
  - Do your results correspond to what is found in Haun et al.?





```{r, eval = FALSE, echo = FALSE}
library(exscidata); library(tidyverse)
data("hypertrophy")


hypertrophy %>%
  select(PARTICIPANT, GROUP, VL_T1, VL_T2, VL_T3, VL_T4) %>%
  pivot_longer(names_to = "time", 
               values_to = "thickness", 
               cols = VL_T1:VL_T4) %>%
  group_by(GROUP, time) %>%
  summarise(m = mean(thickness, na.rm = TRUE), 
            s = sd(thickness, na.rm = TRUE)) 
  



```

