---
title: "qPCR analysis"
author: "Daniel Hammarström"
date: "27 10 2021"
output: html_document
editor_options: 
  chunk_output_type: console
---


## Packages

To import and analyze qPCR data using R, you will need the packages `qpcR`, `parallel` and `qpcrpal`. Installing `qpcrpal` with vignettes will take some time. Set `build_vignettes` to `FALSE` if you are in a hurry.


```{r}

# Check if qpcR, readxl and parallel are installed, otherwise install
if("qpcR" %in% rownames(installed.packages()) == FALSE) install.packages("qpcR")
if("readxl" %in% rownames(installed.packages()) == FALSE) install.packages("readxl")
if("parallel" %in% rownames(installed.packages()) == FALSE) install.packages("parallel")

# Check if qpcrpal is installed, otherwise install
if("qpcrpal" %in% rownames(installed.packages()) == FALSE) {
  
  library(remotes)
  remotes::install_github("dhammarstrom/qpcrpal", build_vignettes = TRUE)
  
}

# Load packages
library(qpcR)
library(readxl)
library(parallel)
library(qpcrpal)
library(tidyverse)





```

## Exploring the data

- Reading and analyzing qPCR-data is done in a stepwise manner using `qpcrpal`.
- The `read_quant5` function organizes the data from the instrument (QuantStudio 5)
- `model_qpcr` creates models of the fluorecense from each reaction
- `analyze_models` use the information in each model to calculate summaries for each reaction.

```{r}


group1 <- read_quant5("./data/qpcr/IDR4000-Group1-precision.xls", skip = 47) %>%
  mutate(ID = paste0(Well, "_", ID))


models <- model_qpcr(group1)


results_group1 <- analyze_models(models)

```


```{r, echo = FALSE, fig.caption = "An example reaction"}

data.frame(predict(models[[1]]), cycle = 1:40) %>%
  ggplot(aes(cycle, Prediction)) +
  geom_line() +
  geom_point(data = group1 %>% filter(ID == "A1_Jacob"),
             aes(Cycle, Rn), shape = 21, fill = "red") +
  labs(title = "Observed and modelled fluorescence per cycle",
       subtitle = "Model: black line; Observed: red points",
       x = "PCR cycle",
       y = "Fluorescence") +
  geom_vline(xintercept = results_group1 %>%
               filter(ID == "A1_Jacob_NA_NA_MyHC2A") %>%
               pull(cpD2),
             lty = 2) +
  annotate("text", x = 17, y = 1.9, label = "Quantification cycle (Cq)",
           hjust = 0) +
  theme_bw()


```


- `qpcrpal` organize the data and uses functions from the `qpcR` package to model the data.
- A quantification cycle can be calculated from each modelled reaction.
- The `analyze_models` function returs alot of values, the cpD2 can be used as the Cq value.

## Exercise: Calculate precision 

- Clone the repository "dhammarstrom/IDR4000-2021" and read a cleaned data set of precision values
- Determine a set of rules (e.g. remove "bad" reactions, the smallest and biggest value, or everything below 10 and above 30)
- Determine the precision for each student as the coefficient of variation $CV = \frac{SD}{Mean}$

- Work in groups: **Who has the lowest CV??**

```{r, eval = FALSE}
precision <- readRDS("./data/rna-lab-cq-precision.RDS")
```



```{r, echo = FALSE}

#
group2 <- read_quant5("./data/qpcr/IDR4000-group2-precision.xls", skip = 47) %>%
  mutate(ID = paste0(Well, "_", ID))
models2 <- model_qpcr(group2)
results_group2 <- analyze_models(models2)


group3 <- read_quant5("./data/qpcr/IDR4000-GROUP3-precision.xls", skip = 47) %>%
  mutate(ID = paste0(Well, "_", ID))

models3 <- model_qpcr(group3)
results_group3 <- analyze_models(models3)


clean_cq_precision <- bind_rows(results_group3  %>%
  separate(ID, into = c("well", "name", "X1", "X2", "target")) %>%
  dplyr::select(name, target, cq = cpD2), 
  results_group2  %>%
  separate(ID, into = c("well", "name", "X1", "X2", "target")) %>%
  dplyr::select(name, target, cq = cpD2), 
  results_group1 %>%
  separate(ID, into = c("well", "name", "X1", "X2", "target")) %>%
  dplyr::select(name, target, cq = cpD2)) 


saveRDS(clean_cq_precision, file = "./data/rna-lab-cq-precision.RDS")

## Example code to generate a table 

# clean_cq_precision%>%
#   filter(cq < 35 & cq > 10) %>%
#   group_by(name, target) %>% 
#   summarise(n = n(), 
#             m = mean(cq), 
#             s = sd(cq), 
#             cv = s / m) %>%
#   arrange(cv) %>%
#   knitr::kable(col.names = c("Name", "Target", "n reactions", "Mean", "SD", "CV"), 
#                digits = c(NA, NA, 0, 2, 2, 3))
# 
```

## Estimation of reaction efficiency

- In theory, a PCR reaction doubles the number of target copies each cycle.
- When plotting log10(concentrations) against Cq values we can estimate efficiency

$$Efficiency = 10^{-1/slope}$$
- When the slope is ~ -3.3

$$Efficiency = 10^{-1/-3.32} = 2$$
- This means that for each cycle we have a doubling of PCR product. A difference in 3.32 cycles would also indicate a 10-fold difference in cDNA concentrations

$$2^{3.32} = 10$$

```{r, echo = FALSE, fig.caption = "Example of a efficiency estimation"}

cyc.diff <- 3.327

dat <- data.frame(concentration = c(1, 1/10, 1/100, 1/1000, 1/10000, 1/100000), 
           cq = c(15, 
                  15 + cyc.diff,
                  15 + cyc.diff*2, 
                  15 + cyc.diff*3, 
                  15 + cyc.diff*4, 
                  15 + cyc.diff*5)) 


slope <- coef(lm(cq ~ log10(concentration), data = dat))[2]

dat %>%
  ggplot(aes(cq, log10(concentration)))  + 
  geom_point() + geom_smooth(method = "lm") + 
  theme_bw() + 
  annotate("text", x = 15, y = -3, hjust = 0, label = "Efficiency factor ~ 2\n when slope ~ 3.3")


```


### Estimation of primer efficiency using real data: Exercises

- Use the data stored in `./data/efficiency-estimate-data.RDS` to plot Cq-values against log10 concentrations. Try to replicate the figure below.
- Create a model (`lm(cq ~ log10(concentration), data = dat`) for each target primer and extract the slope.
- Use the slope to calculate the efficiency factor (`= 10^-(1/slope)`).
- Look at the plot, what is the effect of removing measurements that are clear outliers? Use `filter` to remove specific cq-values from each primer and re-do the calculations.


```{r, echo = FALSE, warning = FALSE, fig.cap = "Efficiency calculations made from serial dilution of cDNA"}

group1 <- read_quant5("./data/qpcr/group1-efficiency.xls", skip = 47) %>%
  mutate(ID = paste0(Well, "_", ID))

models1 <- model_qpcr(group1)

results_group1 <- analyze_models(models1)


eff_g1 <- results_group1 %>%
  separate(ID, into = c("well", "sample", "subsample", "time", "target")) %>%
  dplyr::select(well:target, cq = cpD2) %>%
  filter(!(sample %in% c("FP1", "FP2", "FP3")), 
         target %in% c("reep5", "b2m")) %>%
  mutate(sample = gsub("s1", "", sample), 
         sample = gsub("s2", "", sample), 
         concentration = if_else(sample == "pool1", 10, 
                                 if_else(sample == "pool2", 50, 
                                         if_else(sample == "pool3", 250, 
                                                 if_else(sample == "pool4", 1250, 
                                                         if_else(sample == "pool5", 6250, 
                                                                 if_else(sample == "pool6", 31250, 156250)))))), 
         concentration = 1/concentration) %>%
   dplyr::select(target, concentration, cq) 

#### Group 2

group2 <- read_quant5("./data/qpcr/group2-efficiency.xls", skip = 47) %>%
  mutate(ID = paste0(Well, "_", ID))

models2 <- model_qpcr(group2)

results_group2 <- analyze_models(models2)

eff_g2 <- results_group2 %>%
  separate(ID, into = c("well", "sample", "time", "na", "target")) %>%
  dplyr::select(well:target, cq = cpD2) %>%
  filter(!(sample %in% c("FP1", "FP2", "FP3")), 
         target == "REEP5") %>%
  mutate(concentration = if_else(sample == "P1", 10, 
                                 if_else(sample == "P2", 50, 
                                         if_else(sample == "P3", 250, 
                                                 if_else(sample == "P4", 1250, 
                                                         if_else(sample == "P5", 6250, 
                                                                 if_else(sample == "P6", 31250, 156250)))))), 
         concentration = 1/concentration) %>%
   dplyr::select(target, concentration, cq) 


### Group 3

group3 <- read_quant5("./data/qpcr/group3-efficiency.xls", skip = 47) %>%
  mutate(ID = paste0(Well, "_", ID))

models3 <- model_qpcr(group3)

results_group3 <- analyze_models(models3)

eff_g3 <- results_group3 %>%
  separate(ID, into = c("well", "sample", "conc", "na", "target")) %>%
  dplyr::select(well:target, cq = cpD2) %>%
  filter(!(sample %in% c("fp1", "fp2", "fp3")), 
         target == "B2M") %>%
  mutate(concentration = 1/as.numeric(conc)) %>%
  dplyr::select(target, concentration, cq) 



efficiency_est_data <- bind_rows(eff_g1, eff_g2, eff_g3) %>%
  mutate(target = toupper(target)) %>%
  filter(cq > 10 & cq < 39) 
  
saveRDS(efficiency_est_data, "./data/efficiency-estimate-data.RDS")
  
### Create a figure with all data
 efficiency_est_data %>% 
  ggplot(aes(cq, log10(concentration), fill = target, group = target)) + 
  geom_point(shape = 21, size =  2.5) +
  geom_smooth(method = "lm", se = FALSE, aes(color = target)) +
   labs(x = "Quantification cycle (Cq)", 
        color = "Target", 
        fill = "Target") +
   theme_bw()


## Model efficiency, example solution 
 
 
 
 # Create models and extract the second coefficient from each model (the slope)
b2m_slope <- coef(lm(cq ~ log10(concentration), data = filter(efficiency_est_data, target == "B2M")))[2]

reep5_slope <- coef(lm(cq ~ log10(concentration), data = filter(efficiency_est_data, target == "REEP5")))[2]

# Calculate and store the data in a data frame
efficiency_estimates <- data.frame(target = c("B2M", "REEP5"), 
           Efficiency = c(10^-(1/b2m_slope), 
                          10^-(1/reep5_slope)))


```

## Estimate fold-change

- Use `./data/foldchange-data.RDS` transform Cq-values to a linear scale using an efficiency factor of 2.
- Compute an average normalization reference of all reference genes per sample/timepoint.
- Calculate fold-change of target genes after normalization. 

```{r, echo = FALSE, warning=FALSE}

fc_g1 <- results_group1 %>%
  separate(ID, into = c("well", "sample", "subsample", "time", "target")) %>%
  dplyr::select(well:target, cq = cpD2) %>%
  filter(sample %in% c("FP1", "FP2", "FP3")) %>%
  dplyr::select(sample, time, target, cq) %>%
  mutate(target = toupper(target), 
         target = gsub("RRNA475", "RRNA47S", target)) 



fc_g2 <- results_group2 %>%
  separate(ID, into = c("well", "sample", "time", "na", "target")) %>%
  dplyr::select(well:target, cq = cpD2) %>%
  filter(sample %in% c("FP1", "FP2", "FP3")) %>%
  mutate(target = toupper(target), 
         target = if_else(target %in% c("2A", "2X"), paste0("MYHC", target),
                 if_else(target == "RNA475", "RRNA47S", target))) %>%
  dplyr::select(sample, time, target, cq) 
  

fc_g3 <- results_group3 %>%
  separate(ID, into = c("well", "sample", "time", "target", "na")) %>%
  dplyr::select(well:target, cq = cpD2) %>%
  filter(sample %in% c("fp1", "fp2", "fp3")) %>%
  mutate(sample = toupper(sample), 
         target = toupper(target), 
         time = toupper(time),
         target = if_else(target == "MYCH2A", "MYHC2A", 
                          if_else(target == "MYC2A", "MYHC2A", 
                          if_else(target == "MYCH2X", "MYHC2X", 
                                  if_else(target == "MYC2X", "MYHC2X", 
                                  if_else(target == "MYCH1", "MYHC1", 
                                          if_else(target == "RRNA475", "RRNA47S", target))))))) %>%
    dplyr::select(sample, time, target, cq) 


fc_data <- bind_rows(fc_g1, 
          fc_g2,
          fc_g3)


saveRDS(fc_data, "./data/foldchange-data.RDS")

```







